---
title: "Choosing Resources"
author: "Michael T. Zimmermann"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing Resources}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

RITAN indexes multiple resources and choosing which of them are most relevant for your study can be a challenge. To help with this process, we provide below a set of examples from different types of studies and the thinking behind which resources were used.

# Resources Indexed by default

```{r first_load, echo=TRUE, warning=FALSE, results='hide', message=FALSE}
library(RITANdata)
library(RITAN)
library(knitr)
kable( attr(network_list, 'network_data_sources') )
kable( sapply(geneset_list, length), col.names = c('# Genesets') )
```

While many resources contain information about protein complexes (obligate interactions) and protein-protein interactions (often transient), some use experimental techniques that are specific for physical ineractions. Determining whihc resources indicate physical, through-metabolite, and through-DNA (i.e. transcription factors) interactions, we recommend:
1) each resource's primary publication
2) the [pathguide](http://www.pathguide.org/) website
3) the following guidelines

# Example: Protein Complexes & Protein-Protein Interaction (PPI)

```{r geneset_overlap_P, echo=TRUE, eval=FALSE, warning=FALSE, fig.width = 7, fig.height = 7, fig.align='center'}
genes <- geneset_list$MSigDB_C5$APICAL_JUNCTION_COMPLEX
e <- network_overlap( genes, resources = c('HPRD','CCSB','Biogrid','STRING'), minStringScore = 700 )
## We also strongly encourage use of the BioPlex database, which we do not distribute with RITAN in compliance with their licensing.
```

# Example: Metabolic Network Interaction

```{r geneset_overlap_M, echo=TRUE, eval=FALSE, warning=FALSE, fig.width = 7, fig.height = 7, fig.align='center'}
genes2 <- geneset_list$MSigDB_C5$AMINE_METABOLIC_PROCESS
e2 <- network_overlap( genes, resources = c('PID','HumanNet','Biogrid') )
```

# Example: Expression Signalling

```{r geneset_overlap_E, echo=TRUE, eval=FALSE, warning=FALSE, fig.width = 7, fig.height = 7, fig.align='center'}
genes2 <- geneset_list$MSigDB_C5$AMINE_METABOLIC_PROCESS
e2 <- network_overlap( genes, resources = c('ChEA','HumanNet') )
```


